STAT 672 final project:\\ Stochastic gradient descent

### 2018-04-29 paper/paper.tex
updated paper
wek

\documentclass{article}

@@

\usepackage[citestyle=authoryear]{biblatex}
\addbibresource{paper.bib}

@@

\title{STAT 672 Final Project:\\ Stochastic Gradient Descent}

@@

\section{Introduction}

\subsection{Organization}

@@

\subsection{Motivation}

@@

\subsubsection{Parametric vs. non-parametric}

@@

\subsubsection{Computational Complexity}

@@

\section{Method and Theory}

\subsection{Basic Form}

@@

$f$ takes as input $w \in \mathbb{R}^D$. Assume
that $f$ is convex and differentiable, i.e. we can compute its gradient with
respect to $w$, $\nabla f(w)$. The optimal $w$, i.e. that which minimizes (1),
is denoted $\tilde{w}$. We are willing to settle for a close approximation
$\hat{w}$. The iterative GD algorithm for finding $\hat{w}$ is:

@@

\subsection{Key Properties}

\subsubsection{Correctness}

@@

Now, suppose we use the GD algorithm presented in (2) with $\gamma \leq
\frac{1}{L}$. Denote $x^+ = x - \gamma \nabla
f(x)$ and substitute $x^+$ in for $y$:

@@

Returning to (8), we obtain:

@@

By definition, a squared L2 norm will always be positive unless its content is
0, and we have defined $\gamma$ to be positive, so $\frac{1}{2} \gamma||\nabla f(x)||_2^2$ will always be positive unless the
gradient is equal to zero. So, (9) implies that GD results in a strictly
decreasing objective function value until it reaches the point where the
gradient equals zero, which is the optimal value. A key caveat is an
appropriately chosen $\gamma$, a point covered in more detail later.

The above proof is for GD. Proof is not given for why SGD also
converges to the optimal value. Informally, we can note that the above proof
says given infinite $t$ and appropriate $\gamma$, the iterative algorithm will
always converge to the optimal value. SGD is doing the same thing as GD, 
just with a single observation per iteration 
rather than the entire dataset per iteration. Since SGD is using less information
per iteration, we would expect the convergence to require more iterations. But
per (9) we expect it to \textit{eventually} arrive at the optimal solution.
If so, the difference between GD and SGD must then primarily revolve
around convergence speed, not the final value that is converged to.

\subsubsection{Speed}

Table 1---a simplified version of that in \cite{bottou2010large}---illustrates
the computational advantages of SGD over GD. As shown in row 1, because SGD uses
less information per iteration than GD, it requires more iterations to achieve
fixed degree of accuracy $\rho$.
However, row 2 shows that because GD computes the gradient for every observation every iteration,
while SGD only does so for a single randomly observation, GD's time per
iteration scales linearly with $n$ while SGD's is a constant. Thus, we conclude
that SGD's time to
reach accuracy $\rho$ scales only with $\rho$, while GD's scales with both
$\rho$ and---crucially---linearly with $n$. So, the larger $n$ grows, the larger
SGD's computational advantage over GD.

\begin{table}[h!]
	\centering
	\caption{Asymptotic comparison of GD and SGD}
	\begin{tabular}{|l l l|}
		\hline
		& \textbf{GD} & \textbf{SGD} \\
		Iterations to accuracy $\rho$ & $\log \frac{1}{\rho}$ & $\frac{1}{\rho}$ \\
		Time per iteration & $n$ & 1 \\
		Time to accuracy $\rho$ & $n \log \frac{1}{p}$ & $\frac{1}{\rho}$ \\
		\hline
	\end{tabular}
\end{table}

Figure 1 visually compares SGD's convergence to that of GD (and
mini-batch gradient descent, a variation addressed later in this
paper).\footnote{Image taken from www.towardsdatascience.com} The
path of SGD's convergence is very squiggly, reflecting the extra variation
introduced by only using a single observation per iteration. The path of GD's
convergence is much smoother, reflecting the use of all observations per
iteration.

\begin{figure}[h!]
	\centering
	\caption{Convergence}
\includegraphics[scale=0.3]{img}
\end{figure}

\subsection{Extensions}

@@

\subsubsection{Step Size}

@@

\subsubsection{Momentum and Acceleration}

@@

\subsubsection{Averaging}

@@

\subsubsection{Mini-batch}

@@

\subsubsection{Parallelization}

@@

\section{Applications}

@@

\subsection{SGD and Statistical Learning}

Consider a typical supervised classification problem. We have
feature matrix $\bm{X}_{n \times D}$ ($n$ observations and $D$
features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
predict an observation's label $Y_i$ using that observation's features
$\bm{X}_i$. To emphasize the utility of SGD, suppose that both $n$ and $D$ are
very large.
We have a hypothesis class $\mathcal{F}$ of unctions $f(\bm{w}, \bm{X}_i) \in \mathcal{F}$ parametrized by weight
vector $\bm{w}$. We have convex loss function $L(Y_i, f(\bm{w}, \bm{X}_i))$ that
expresses the cost of misclassification. 
We will consider the optimal function
that which minimizes empirical risk over all observations: $\frac{1}{n}
\sum_{i=1}^n L(Y_i, f(\bm{w}, \bm{X}_i))$.
Denote $\hat{\bm{w}}$ the weight coefficients of this optimal function. We thus
have:

@@

\subsubsection{Support vector machines}

@@

There have been further efforts to improve SGD for the specific application of
SVM, most notably \cite{shalev2011pegasos}.

\subsection{Simulation Case Study}

Simulation provides an example of the computational advantages of SGD. 
Suppose we have one-dimensional feature matrix $\bm{X}$ with $X_i \sim N(0, 1)$ 
and outcome $\bm{Y}$ with the
relationship $Y_i \sim (0.5 X_i) + \epsilon$. There are 10 million observations.

\begin{lstlisting}[language=Python]
import numpy as np
X = np.random.randn(int(10e6))
Y = (0.5 * X) + np.randn(1)
\end{lstlisting}

We seek to estimate coefficients in a regression $y = \beta_0 + \beta_1 X$ by least
squares. We do so via two competing methods: batch GD and SGD.

\begin{lstlisting}[language=Python]
def GD(X, Y, gamma, epsilon, maxiter):
	n = Y.size
	w = 1 #initial guess
	for i in range(maxiter):
		Yhat = np.dot(X, w)
		update = w - gamma * (np.dot(X.T, Yhat - Y) / n)
		if (abs(w - update) < epsilon): 
			break
		else:
			w = update
	return(w)
\end{lstlisting}
\begin{lstlisting}[language=Python]
def SGD(X, Y, gamma, epsilon, maxiter):
	n = Y.size
	w = 1 #initial guess
	for i in range(maxiter):
		j = np.random.randint(0, n)
		Yhat = w * X[j]
		update = w - gamma * ((Yhat - Y[j])*X[j])
		if (abs(w - update) < epsilon):
			break
		else:
			w = update
	return(w)
\end{lstlisting}

Both algorithms were run on the $n=10^6$ dataset with the same step size
$\gamma$ and precision $\epsilon$. As shown in Table 2, both algorithms 
converge to essentially the correct answer, but SGD does so in dramatically less time than GD.

@@

\subsection{Real-World Applications}

@@

\textbf{ImageNet} is a dataset consisting of millions
of labeled images. The
		ImageNet Large Scale Visual Recognition Competition (ILSVRC) is
		an annual contest in which competing teams train models for
		automatic image classification. The submission of \cite{krizhevsky2012imagenet} used convolutional neural
		nets to achieve dramatic reduction in error rates and is
		one of the most cited machine learning papers of all time. Per the aforementioned paper: ``We
		trained our models using stochastic gradient descent with a
		batch size of 128 examples, momentum of 0.9, and weight decay of
		0.00005.''

Google's \textbf{AlphaGo} software applies deep learning to the
		board game of Go and has received extensive attention for its
		ability to defeat even top human players (e.g., twice making the
		cover of \textit{Nature} with \cite{silver2016mastering} and
		\cite{silver2017mastering}). Google's slides from the 2016
		International Conference on Machine Learning (ICML) indicate
		that SGD is used in every stage of the
		model: supervised learning of policy networks, reinforcement
		learning of policy networks, and reinforcement learning of value
		networks.

The \textbf{Netflix Grand Prize} was a competition hosted by
		Netflix to create better movie recommendation models
		(\cite{bennett2007netflix}). Most of the top entries used
		SGD for estimating model coefficients; for example,
		\cite{koren2009bellkor} states that ``in order to learn the
		involved parameters$\ldots$ learning is done by a stochastic
		gradient descent algorithm running for 30 iterations.''

\section{Conclusion}

>>>>

\documentclass{report}

@@

\usepackage[citestyle=authoryear, style=authoryear]{biblatex}
\addbibresource{paper.bib}

@@

\title{STAT 672 final project:\\ Stochastic gradient descent}

@@

\tableofcontents

\chapter{Introduction}

\section{Organization}

@@

\section{Motivation}

@@

\subsection{Parametric vs. non-parametric}

@@

\subsection{Computational complexity}

@@

\chapter{Method and theory}

\section{Basic form}

@@

$f$ takes as input $w \in \mathbb{R}^D$. Assume
that $f$ is convex and differentiable, i.e. we can compute its gradient with
respect to $w$, $\nabla f(w)$. The optimal $w$, i.e. that which minimizes (2.1),
is denoted $\tilde{w}$. We are willing to settle for a close approximation
$\hat{w}$. The iterative GD algorithm for finding $\hat{w}$ is:

@@

\section{Key properties}

\subsection{Correctness}

@@

Now, suppose we use the GD algorithm presented in (2.2) with $\gamma \leq
\frac{1}{L}$. Denote $x^+ = x - \gamma \nabla
f(x)$ and substitute $x^+$ in for $y$:

@@

Returning to (2.8), we obtain:

@@

A squared L2 norm will always be positive unless its content is
0, and we have defined $\gamma$ to be positive, so $\frac{1}{2} \gamma||\nabla f(x)||_2^2$ will always be positive unless the
gradient is equal to zero. So, (2.9) implies that GD results in a strictly
decreasing objective function value until it reaches the point where the
gradient equals zero, which is the optimal value. A key caveat is an
appropriately chosen $\gamma$, a point covered in more detail later.

The above proof is for GD. Proof is not given for why SGD also
converges to the optimal value. Informally, we can note that the above proof
says given infinite $t$ and appropriate $\gamma$, the iterative algorithm will
always converge to the optimal value. SGD is doing the same thing as GD, 
just with a single observation per iteration 
rather than the entire dataset per iteration. Since SGD is using less information
per iteration, we would expect the convergence to require more iterations. But
per (2.9) we expect it to \textit{eventually} arrive at the optimal solution.
If so, the difference between GD and SGD must then primarily revolve
around convergence speed, not the final value that is converged to.

\subsection{Speed}

Table 2.1---a simplified version of that in \cite{bottou2010large}---illustrates
the computational advantages of SGD over GD. As shown in row 1, because SGD uses
less information per iteration than GD, it requires more iterations to achieve
fixed degree of accuracy $\rho$.
However, row 2 shows that because GD computes the gradient for every observation every iteration,
while SGD only does so for a single randomly observation, GD's time per
iteration scales linearly with $n$ while SGD's is a constant. Thus, we conclude
that SGD's time to
reach accuracy $\rho$ scales only with $\rho$, while GD's scales with both
$\rho$ and---crucially---linearly with $n$. So, the larger $n$ grows, the larger
SGD's computational advantage over GD.

\begin{table}[t]
	\centering
	\caption{Asymptotic comparison of GD and SGD}
	\begin{tabular}{|l l l|}
		\hline
		& \textbf{GD} & \textbf{SGD} \\
		Iterations to accuracy $\rho$ & $\log \frac{1}{\rho}$ & $\frac{1}{\rho}$ \\
		Time per iteration & $n$ & 1 \\
		Time to accuracy $\rho$ & $n \log \frac{1}{p}$ & $\frac{1}{\rho}$ \\
		\hline
	\end{tabular}
\end{table}

Figure 2.1 visually compares SGD's convergence to that of GD (and
mini-batch gradient descent, a variation addressed later in this
paper).\footnote{Image taken from www.towardsdatascience.com} The
path of SGD's convergence is very squiggly, reflecting the extra variation
introduced by only using a single observation per iteration. The path of GD's
convergence is much smoother, reflecting the use of all observations per
iteration.

\begin{figure}[t]
	\centering
	\caption{Convergence}
\includegraphics[scale=0.3]{img}
\end{figure}

\section{Extensions}

@@

\subsection{Step size}

@@

\subsection{Momentum and acceleration}

@@

\subsection{Averaging}

@@

\subsection{Mini-batch}

@@

\subsection{Parallelization}

@@

\chapter{Applications}

@@

\section{SGD and statistical learning}

Consider a typical supervised classification problem. We have
feature matrix $\bm{X}_{n \times D}$ ($n$ observations and $D$
features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
predict an observation's label $Y_i$ using that observation's features
$\bm{X}_i$. To emphasize the utility of SGD, suppose that both $n$ and $D$ are
very large.
We have a hypothesis class $\mathcal{F}$ of functions $f(\bm{w}, \bm{X}_i) \in \mathcal{F}$ parametrized by weight
vector $\bm{w}$. We have convex loss function $L(Y_i, f(\bm{w}, \bm{X}_i))$ that
expresses the cost of misclassification. 
We will consider the optimal function
that which minimizes empirical risk over all observations: $\frac{1}{n}
\sum_{i=1}^n L(Y_i, f(\bm{w}, \bm{X}_i))$.
Denote $\hat{\bm{w}}$ the weight coefficients of this optimal function. We thus
have:

@@

\subsection{Support vector machines}

@@

There have been further efforts to improve SGD for the specific application of
SVM, most notably the Pegasos algorithm of \cite{shalev2011pegasos}.

\section{Simulation case study}

Simulation provides an example of the computational advantages of SGD. 
Suppose we have one-dimensional feature matrix $\bm{X}$ with $X_i \sim N(0, 1)$ 
and outcome $\bm{Y}$ with the
relationship $Y_i \sim (0.5 X_i) + \epsilon$. There are 10 million observations.
We seek to estimate coefficients in a regression $y = \beta_0 + \beta_1 X$ by least
squares. We do so via two competing methods: batch GD and SGD.
Both algorithms were run on the $n=10^6$ dataset with the same step size
$\gamma$ and precision $\epsilon$. As shown in Table 3.1, both algorithms 
converge to essentially the correct answer, but SGD does so in dramatically less time than GD.

\begin{figure}[h!]
	\caption{Generating data}
\begin{lstlisting}[language=Python]
import numpy as np
X = np.random.randn(int(10e6))
Y = (0.5 * X) + np.randn(1)
\end{lstlisting}
\end{figure}

\begin{figure}[h!]
	\caption{Batch GD algorithm}
	\begin{lstlisting}[language=Python]
	def GD(X, Y, gamma, epsilon, maxiter):
	n = Y.size
	w = 1 #initial guess
	for i in range(maxiter):
		Yhat = np.dot(X, w)
		update = w - gamma * (np.dot(X.T, Yhat - Y) / n)
		if (abs(w - update) < epsilon): 
			break
		else:
			w = update
	return(w)
	\end{lstlisting}
\end{figure}

\begin{figure}[h!]
	\caption{SGD algorithm}
\begin{lstlisting}[language=Python]
def SGD(X, Y, gamma, epsilon, maxiter):
	n = Y.size
	w = 1 #initial guess
	for i in range(maxiter):
		j = np.random.randint(0, n)
		Yhat = w * X[j]
		update = w - gamma * ((Yhat - Y[j])*X[j])
		if (abs(w - update) < epsilon):
			break
		else:
			w = update
	return(w)
\end{lstlisting}
\end{figure}

@@

\section{Real-world applications}

@@

\subsection{ImageNet}
ImageNet is a dataset consisting of millions
of labeled images. The
		ImageNet Large Scale Visual Recognition Competition (ILSVRC) is
		an annual contest in which competing teams train models for
		automatic image classification. The submission of \cite{krizhevsky2012imagenet} used convolutional neural
		nets to achieve dramatic reduction in error rates and is
		one of the most cited machine learning papers of all time. Per the aforementioned paper: ``We
		trained our models using stochastic gradient descent with a
		batch size of 128 examples, momentum of 0.9, and weight decay of
		0.00005.''
\subsection{AlphaGo}
Google's AlphaGo software applies deep learning to the
		board game of Go and has received extensive attention for its
		ability to defeat even top human players (e.g., twice making the
		cover of \textit{Nature} with \cite{silver2016mastering} and
		\cite{silver2017mastering}). Google's slides from the 2016
		International Conference on Machine Learning (ICML) indicate
		that SGD is used in every stage of the
		model: supervised learning of policy networks, reinforcement
		learning of policy networks, and reinforcement learning of value
		networks.

\subsection{Netflix grand prize}
The Netflix Grand Prize was a competition hosted by
		Netflix to create better movie recommendation models
		(\cite{bennett2007netflix}). Most of the top entries used
		SGD for estimating model coefficients; for example,
		\cite{koren2009bellkor} states that ``in order to learn the
		involved parameters$\ldots$ learning is done by a stochastic
		gradient descent algorithm running for 30 iterations.''

\chapter{Conclusion}
